---
title: "Class 11 Lab"
author: "Andres Sandoval"
format: gfm
---


# Section 1 - Proportion of G/G in a population 

#### MXL - Mexican Ancestry in Los Angeles

Downloaded csv file from Ensemble. 

Here we read the csv file. 

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

head(mxl)
```

```{r}
table( mxl$Genotype..forward.strand. )
#table function summarizes the data in the table
```


```{r}
round(table(mxl$Genotype..forward.strand.) / nrow(mxl) *100, 2)
```

Let's try examining another population. 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G in GBR 

```{r}
round(table(gbr$Genotype..forward.strand.)/ nrow(gbr) *100, 2 )
```
This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population. 

Let's now dig into this further. 


# Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 


>Q13

How many samples do you have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```


How many individuals?
```{r}
nrow(expr)
```

How many samples of each genotype?
```{r}
table(expr$geno)
```


#### BoxPlot section

```{r}
library(ggplot2)
```

Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3 expression.

Let's make our boxplot with this data. 

```{r}
boxplot <- ggplot(expr) + aes(x=geno, y = exp, fill= geno) + geom_boxplot(notch = TRUE)
```

>Q13. What is the median expression of each of the genotypes?

```{r}
library(dplyr)
expr %>% group_by(geno) %>% summarize(median = median(exp))
```


>Q14. Boxplot

```{r}
boxplot
```



