---
title: "Class 10: Mini Project"
author: "Andres Sandoval"
format: pdf
---

## Importing Candy Data

```{r}
candy_file <- "candy-data.txt"
candy = read.csv(candy_file, row.names = 1)
head(candy)
```


>Q1. How many different candy types are in this dataset?

```{r}
dim(candy)
nrow(candy)
```
There are 85 different candy types in this dataset.



>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy[,"fruity"])
table(candy$fruity)
```
There are 38 fruity candy types in the data set. 


## What is your favorite candy

```{r}
candy["Twix", ]$winpercent
```



>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
#first view all the names of candies
row.names(candy)
```
```{r}
#view favorite candy and its win percent
round(candy["Skittles original",]$winpercent, 2)
```

Skittles original has a win percent of 63.09%

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
round(candy["Kit Kat",]$winpercent, 2)
```


>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
round(candy["Tootsie Roll Snack Bars",]$winpercent, 2)
```



>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
library(skimr)
skim(candy)
```

The "winpercent" column seems to be on a different scale



>Q7. What do you think a zero and one represent for the candy$chocolate column?
Hint: look at the “Variable type” print out from the skim() function. Most varables (i.e. columns) are on the zero to one scale but not all. Some columns such as chocolate are exclusively either zero or one values.


```{r}
skim(candy$chocolate)
```

A value of 0 indicates that there is no chocolate within the candy, while a value of 1 indicates that there is chocolate in the candy. 


>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```



>Q9. Is the distribution of winpercent values symmetrical?

The distribution is not symmetrical. 

>Q10. Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)
```
The center of the distribution is below 50 %


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}

mean (candy$winpercent[as.logical(candy$chocolate)] )
mean (candy$winpercent[as.logical(candy$fruity)] )

```
On average, chocolate candy is higher ranked than fruity candy. 

>Q12. Is this difference statistically significant?

```{r}
t.test( candy$winpercent[as.logical(candy$chocolate)],  candy$winpercent[as.logical(candy$fruity)] )
```


Due to the p-value of 2.87e-08, we can reject the null hypothesis. Therefore, we have enough evidence to declare the two means are different. 



## Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)

library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```
The five least liked candies are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

>Which apprach do you prefer and why?

I prefer the `order()` function of R studio as it does not require the dplyr package. 

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5)

```

The top five are Reese's Peanity Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers. 



>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```


>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?


```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col() 
```

*Time to add some useful color*

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
  
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```



>Q17. What is the worst ranked chocolate candy?

Sixlets




>Q18. What is the best ranked fruity candy?

Starburst






# Taking a Look at Pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 40)
```




>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?


The highest ranked is fruity type candy. 


>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?


```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )

```
The least popular would be Strawberry bon bons. 





>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy),pricepercent)) +
  geom_col(fill=my_cols)
```

```{r}
#Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col=my_cols) +
    geom_point()
```

## Exploring the Correlation Structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```


>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity 



>Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar 




##  Principal Component Analysis

Let’s apply PCA using the `prcomp()` function to our candy dataset remembering to set the scale=TRUE argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```


```{r}
pca$rotation[,1]
```

Now plot the results of PC1 vs PC2

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```



```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

# Make a ggplot
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```


```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 40)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")

```


```{r}
library(plotly)
```

```{r}
#ggplotly(p)
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity and hard. This makes sense as these two values maintained a a relatively high positive correlation with one another. 


